Running an open experiment: transparency and reproducibility in soil and ecosystem

### 2016-06-28 manuscript_openexperiment/Response_to_reviewers.Rmd

---
title: "Response_to_reviewers"
author: "Ben Bond-Lamberty"
date: "June 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 2016-06-28 manuscript_openexperiment/Response_to_reviewers.Rmd

---
title: "Response_to_reviewers"
author: "Ben Bond-Lamberty"
date: "June 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Referee 2**

*This is a rather unusual manuscript, reporting not on science but on a new way that science might be conducted.   I was prepared to dislike the paper, but I found it rather interesting and easy to read. It might even convince me to join a disaggregate group of scientists to pursue a study of interest.  If there is a special section of ERL that publishes such papers, I would encourage you to include it.*

Thank you.

*Only a couple of specialized comments*
*1.       The authors should standardize whether web citations are given in the text or in footnotes.  I prefer the latter to keep the text clean.*

We agree, and have moved all web citations--except the one referencing the main repository, the subject of this manuscript--to footnotes. (We will obviously follow journal requirements if they differ from this, however.)

*2.       Figure 2, particularly the lower panel, needs a lot more explanation before it will be easy for readers to understand.*

We have expanded and hopefully clarified the Figure 2 caption.

**Referee 3**

*The manuscript is a perspective and case study analysis on the value of running an "open" experiment. The idea behind this is to lay out the protocol, the treatments, the results and even the manuscript drafts all open for public viewing and commenting in real time, as data come in and results are developed. It's a fresh and radical approach to doing science and could strongly benefit rebuilding trust in our field, as the authors document and adequately reference literature on the topic.*

*As a journal editor, I have been pushing in every article to provide a clear data access statement and to publish data in open and public repositories. It's nice to have a manuscript to point to on good reason for doing so with a clear example.*

Thank you.

*I don't have any recommendations for major changes to this short piece. However, I think the caveats section (section 3) could benefit from more discussion on barriers to adopting this approach, not just from the incentive side, but from the skills needed, time invested, and challenges in using systems like github to run experiments. It would be helpful to discuss what was hard in getting the open experiment to work, what took the most time, what tools were the most difficult or least developed or accessible to scientists. If anything most of us are well convinced of the arguments the authors make on why to share data and code and incentives are showing up in multiple places. The challenge is finding the time, motivation, and technical skills to do so. So perhaps some discussion of how the authors dealt with those would help convince the otherwise skittish researcher that this isn’t as hard as it looks to do!*

TODO.

### 2016-06-28 manuscript_openexperiment/Open_experiment.Rmd
Responses to Reviewer 2.

---
title: 'Running an open experiment: transparency and reproducibility in soil and ecosystem
  science'
author: "Ben Bond-Lamberty^1^, A. Peyton Smith^2^, Vanessa Bailey^2^"
date: "February 2, 2016"
output: word_document
---
^1^Joint Global Change Research Institute, DOE Pacific Northwest National Laboratory, College Park, MD USA; 
bondlamberty@pnnl.gov

@@

As data and code have become increasingly intertwined, the availability of the latter has become a fundamental problem as well. Specialized modeling groups have worked to improve reproducibility and archiving practices {Thornton, 2005 #2310}. More generally, scientists in all fields are increasingly building and using software in their work, though often without strong training in this area {Wilson, 2014 #4337}. In addition to open-source data analysis languages such as R {R Development Core Team, 2016 #4314}, scientific workflow systems such as Kepler (https://kepler-project.org) or Taverna (http://www.taverna.org.uk) record information about the data processing, analytical process and decisions, and statistical analysis. Providing open code does not magically produce bug-free code, mistake-free analyses, or instantly better science {Easterbrook, 2014 #4137}. But it does encourage authors to invest the time upfront to clean up their code, data, and documentation when a paper is written, rather than deferring this task--often until key details have been forgotten, if not forever. This also allows for real-time peer review of both code and data.

@@

* The entire data processing and analytical system is online and documented. It is written in R {R Development Core Team, 2016 #4314}, an open-source and widely used language and environment for statistical computing and graphics.
* The version control system let us make incremental changes, work out problems, look at histories (i.e., who made what change when).
* An issue tracker let us discuss problems, reference changes in the repository, create to-do lists, and assign responsibilities.
* An informational webpage provided non-technical explanations of the experiment and broader project.
* Manuscripts were directly tied into the data system (cf. **Figure 1**), with numerical results flowing directly into e.g. results sections. This meant that changes to the data (and thus statistical results, etc.) propagated automatically and consistently. Tools such as R Markdown[^7] have made this process much easier to build and maintain.
* Log files provide an audit trail {Ellison, 2010 #4363} of what analytical steps were taken and, critically, the specific versions used for each unit of software, including the main R system itself. This is critical as software changes with time, a potentially large problem in reproducing previous analyses.

[^6]: http://rmarkdown.rstudio.com

@@

The model we present here--a public repository updated throughout the experiment, analysis, and publication process--is far from perfect. First, there are a number of "best practices" of scientific computing {Wilson, 2014 #4337} that we did not employ, in particular with respect to automation. Second, this approach may not be appropriate for, nor applicable to, all study types, even in our professional fields of soil, ecosystem, and global change science; it certainly is not appropriate for sensitive human-subject data. Third, Git and GitHub are designed for working with code, not data. Although this situation continues to improve on a technical level[^7], it still requires a fair amount of coding expertise to set up and run an experimental repository such as the one detailed here. We applaud efforts to lower the technical and information hurdles to open science {Hampton, 2015 #4420}.

[^7]: https://github.com/blog/1885-better-word-highlighting-in-diffs

@@

**Figure 2.** Two of the many diagnostic figures that were produced as the incubation experiment proceeded. They show (top) the number of replicate measurements made on each soil core, by date of measurement, and (bottom) measurements with a data mismatch problem, indicating a data entry or other error (in the figure no errors, which would be in red, are extant). A wide range of similar diagnostic graphics (see https://github.com/bpbond/cpcrw_incubation#current-diagnostics) and statistics let us quickly identify problems or interesting patterns in the data. All diagnostics were publicly available in real time.

>>>>

---
title: 'Running an open experiment: transparency and reproducibility in soil and ecosystem
  science'
author: "Ben Bond-Lamberty^1^, A. Peyton Smith^2^, Vanessa Bailey^2^"
date: "July 1, 2016"
output:
  html_document: default
  word_document: default
---
^1^Joint Global Change Research Institute, DOE Pacific Northwest National Laboratory, College Park, MD USA; 
bondlamberty@pnnl.gov

@@

As data and code have become increasingly intertwined, the availability of the latter has become a fundamental problem as well. Specialized modeling groups have worked to improve reproducibility and archiving practices {Thornton, 2005 #2310}. More generally, scientists in all fields are increasingly building and using software in their work, though often without strong training in this area {Wilson, 2014 #4337}. In addition to open-source data analysis languages such as R {R Development Core Team, 2016 #4314}, scientific workflow systems such as Kepler[^6] or Taverna[^7] record information about the data processing, analytical process and decisions, and statistical analysis. Providing open code does not magically produce bug-free code, mistake-free analyses, or instantly better science {Easterbrook, 2014 #4137}. But it does encourage authors to invest the time upfront to clean up their code, data, and documentation when a paper is written, rather than deferring this task--often until key details have been forgotten, if not forever. This also allows for real-time peer review of both code and data.

[^6]: https://kepler-project.org
[^7]: http://www.taverna.org.uk

@@

* The entire data processing and analytical system is online and documented. It is written in R {R Development Core Team, 2016 #4314}, an open-source and widely used language and environment for statistical computing and graphics.
* The version control system let us make incremental changes, work out problems, look at histories (i.e., who made what change when).
* An issue tracker let us discuss problems, reference changes in the repository, create to-do lists, and assign responsibilities.
* An informational webpage provided non-technical explanations of the experiment and broader project.
* Manuscripts were directly tied into the data system (cf. **Figure 1**), with numerical results flowing directly into e.g. results sections. This meant that changes to the data (and thus statistical results, etc.) propagated automatically and consistently. Tools such as R Markdown[^8] have made this process much easier to build and maintain.
* Log files provide an audit trail {Ellison, 2010 #4363} of what analytical steps were taken and, critically, the specific versions used for each unit of software, including the main R system itself. This is critical as software changes with time, a potentially large problem in reproducing previous analyses.

[^8]: http://rmarkdown.rstudio.com

@@

The model we present here--a public repository updated throughout the experiment, analysis, and publication process--is far from perfect. First, there are a number of "best practices" of scientific computing {Wilson, 2014 #4337} that we did not employ, in particular with respect to automation. Second, this approach may not be appropriate for, nor applicable to, all study types, even in our professional fields of soil, ecosystem, and global change science; it certainly is not appropriate for sensitive human-subject data. Third, Git and GitHub are designed for working with code, not data. Although this situation continues to improve on a technical level[^9], it still requires a fair amount of coding expertise to set up and run an experimental repository such as the one detailed here. We applaud efforts to lower the technical and information hurdles to open science {Hampton, 2015 #4420}.

[^9]: https://github.com/blog/1885-better-word-highlighting-in-diffs

@@

**Figure 2.** Two of the many diagnostic figures that were produced as the incubation experiment proceeded. Top: the number of replicate measurements made on each soil core (y), by date of measurement (x); the 'gap' at 2015-11-11/12 shows a potential problem in our data-processing pipeline, related to the instrument's clock crossing midnight. Bottom: data matches between multiplexer valve number (y) and soil core, by date of measurement (x); grey dots indicate that every valve has been correctly matched with a soil core, while a red dot would indicate a data entry problem or other error. A wide range of similar diagnostic graphics and statistics let us quickly identify problems or interesting patterns in the data. All diagnostics were publicly available in real time.
